Retrofitting Effect Handlers onto OCaml

### 2021-03-30 retro-concurrency.tex
updates
cleanup

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect In_line : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |In_line|, which is parameterised with an input
channel of type |in_channel|, which when \emph{performed} returns a |string|
value. A computation can perform the |In_line| effect without knowing how the
|In_line| effect is implemented. This computation may be enclosed by different
handlers that handle |In_line| differently. For example, |In_line| may be
implemented by performing a blocking read on the input channel or performing
the read asynchronously by offloading it to an event loop such as |libuv|,
without changing the computation. Thanks to the separation of effectful
operations from their implementation, effect handlers enable new approaches to
modular programming. Effect handlers are a generalisation of exception
handlers, where, in addition to the effect being handled, the handler is
provided with the delimited continuation~\cite{Danvy90} of the perform site,
which may be resumed later. This enables non-local control-flow mechanisms such
as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed.

@@

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Hillerstrom20,Pyro,Frank,Eff}. Unlike these efforts,
our goal is to retrofit effect handlers onto the OCaml programming language,
which has been in continuous use for the past 25 years in large codebases
including verification tools~\cite{everest,Coq}, mission critical software
systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with a fast foreign-function interface
(FFI). It has excellent compatibility with program analysis tools such as
debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
obtain a backtrace.

@@

\subsection{Requirements}
\label{sec:req}

@@

\begin{itemize}
	\item the design and implementation of effect handlers for OCaml. Our design
		retains OCaml's compatibility with program analysis tools that inspect the
		stack using DWARF unwind tables. We have validated our DWARF unwind tables
		with the assistance of an automated validator tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the subtle interactions with the C
		stack, which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has minimal
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

The implementation follows the operational semantics, but has a few key
representational differences. Unlike the operational semantics, handlers with
just exception patterns (exception handlers) are implemented differently than
effect handlers. As mentioned in \S\ref{sec:exn_handlers}, exceptions are
pervasive in OCaml and are so cheap that they are used for local control flow.
Hence, we retain the linked exception handler frame implementation of stock
OCaml exception handlers in Multicore OCaml for performance backwards
compatibility. This differs from other research languages with effect
handlers~\cite{Hillerstrom20,Frank,Eff}, which implement exceptions using
effects (by ignoring the continuation argument in the handler).

@@

Similar to exception handlers, the lack of callee-saved registers in OCaml
benefits effect handlers. There is no register state to save when entering an
effect handler or performing an effect. Similarly, there is no register state
to restore when handling an effect or resuming a continuation. This fortuitous
design choice in stock OCaml has a significant impact towards fast switching
between fibers in Multicore OCaml.

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. This bytecode must
be interpreted on demand from the start of the function definition to compute
the call-frame information at any given instruction within the function. For
each function, DWARF maintains a \emph{canonical frame address} (CFA) and is
traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|.

Our goal is to encode the CFA using DWARF directives when stacks are switched.
Recall that stack switching occurs in effect handlers, external calls and
callbacks. At the entry to a effect handler block, we insert DWARF bytecode to
follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
the C stack. DWARF unwinding for external calls is implemented by following a
link to the current OCaml stack pointer. With these changes, we get the same
backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
runtime system functions due to effect handlers. We have verified the
correctness of our DWARF directives using the verification tool from Bastian et
al.~\cite{Bastian19}.

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly, which speeds up DWARF-based unwinding by up to 25$\times$. With this
technique, |perf| can unwind the stack at sample points rather than dumping the
call stack, which captures the complete backtrace rather than just the current
fiber.

\vspace{-2mm}
\subsection{Garbage collection}
\label{sec:gc}

@@

For stack walking during GC in order to find the roots, the only change
necessary was to make the runtime aware of the fiber layout and how to follow
the parent pointers. The stack map generation and scanning each frame for roots
remains the same. The challenges and the solutions for integrating fibers with
the concurrent mark-and-sweep GC of Multicore OCaml have been discussed
previously~\cite{Sivaramakrishnan20}.

\vspace{-2mm}
\section{Evaluation}
\label{sec:eval}

In this section, we evaluate the performance of Multicore OCaml effect handlers
against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
an extension of the 4.10.0 OCaml compiler with support for shared memory
parallelism and effect handlers. Since our objective is to evaluate the impact
of effect handlers, none of our benchmarks utilise parallelism. These results
were obtained on a 2-socket Intel\textregistered Xeon\textregistered Gold 5120
x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of
main memory.

\vspace{-2mm}
\subsection{No effects benchmarks}

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation
of Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation
and re-allocation of resources every time the non-local control leaves and
enters back into a context. |dynamic-wind| is not quite the right abstraction
as resources need to be cleaned up only on non-returning
exists~\cite{Kiselyov,Sitaram03}. This requires distinguishing returning exits
from non-returning ones.

>>>>

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect In_line : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |In_line|, which is parameterised with an input
channel of type |in_channel|, which when \emph{performed} returns a |string|
value. A computation can perform the |In_line| effect without knowing how the
|In_line| effect is implemented. This computation may be enclosed by different
handlers that handle |In_line| differently. For example, |In_line| may be
implemented by performing a blocking read on the input channel or performing
the read asynchronously by offloading it to an event loop such as |libuv|,
without changing the computation. Thanks to the separation of effectful
operations from their implementation, effect handlers enable new approaches to
modular programming. Effect handlers are a generalisation of exception
handlers, where, in addition to the effect being handled, the handler is
provided with the delimited continuation~\cite{Danvy90} of the perform site.
This continuation may be used to resume the suspended computation later. This
enables non-local control-flow mechanisms such as resumable exceptions,
lightweight threads, coroutines, generators and asynchronous I/O to be
composably expressed.

@@

There are many research languages and libraries built around effect
handlers~\cite{Leijen14,Biernacki20,Hillerstrom20,Pyro,Frank,Eff}. Unlike these
efforts, our goal is to retrofit effect handlers onto the OCaml programming
language, which has been in continuous use for the past 25 years in large
codebases including verification tools~\cite{everest,Coq}, mission critical
software systems~\cite{astree} and latency sensitive networked
applications~\cite{Madhavapeddy13}. OCaml is particularly favoured for its
competitive yet predictable performance, with a fast foreign-function interface
(FFI). It has excellent compatibility with program analysis tools such as
debuggers and profilers that utilise DWARF stack unwind tables~\cite{DWARF} to
obtain a backtrace.

@@

\vspace{-1mm}
\subsection{Requirements}
\label{sec:req}

@@

\begin{itemize}
	\item the design and implementation of effect handlers for OCaml. Our design
		retains OCaml's compatibility with program analysis tools that inspect the
		stack using DWARF unwind tables. We have validated our DWARF unwind tables
		with the assistance of an automated validator tool~\cite{Bastian19}.
	\item a formal operational semantics for the effect handler implementation in
		OCaml. Our formalism explicitly models the interactions with the C stack,
		which is generally overlooked by other formal models, but which the
		implementations must handle.
	\item extensive evaluation which shows that our implementation has minimal
		impact on code that does not use effect handlers, and serves as an
		efficient foundation for scalable concurrent programming.
\end{itemize}

@@

The implementation follows the operational semantics, but has a few key
representational differences. Unlike the operational semantics, handlers with
just exception patterns (exception handlers) are implemented differently than
effect handlers. As mentioned in \S\ref{sec:exn_handlers}, exceptions are
pervasive in OCaml and are so cheap that they are used for local control flow.
Hence, we retain the linked exception handler frame implementation of stock
OCaml in Multicore OCaml to ensure performance backwards compatibility. This
differs from other research languages with effect
handlers~\cite{Hillerstrom20,Frank,Eff}, which implement exceptions using
effects (by ignoring the continuation argument in the handler).

@@

Similar to exception handlers, the lack of callee-saved registers in OCaml
benefits effect handlers. There is no register state to save when entering an
effect handler or performing an effect. Similarly, there is no register state
to restore when handling an effect or resuming a continuation. This fortuitous
design choice in stock OCaml has a significant impact in enabling fast
switching between fibers in Multicore OCaml.

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation that describes the
unwind table as a sequence of edits from the start of the function. In order to
compute the call-frame information at any given instruction within a function,
the DWARF bytecode from the start of that function must be interpreted on
demand. For each function, DWARF maintains a \emph{canonical frame address}
(CFA) and is traditionally the stack pointer before entering this function.
Hence, on x86-64, where the return address is pushed on the stack on call, the
return address is at |CFA - 8|.

Our goal is to compute the CFA of the caller when stacks are switched using the
DWARF directives. Recall that stack switching occurs in effect handlers,
external calls and callbacks. At the entry to an effect handler block, we insert
DWARF bytecode to follow the |parent_fiber| pointer and dereference the
|saved_sp| to get the CFA (|saved_sp + 8|). During callbacks into OCaml, we
save the current system stack pointer in the |context block| in
Figure~\ref{fig:fiber} to identify the CFA in the C stack. DWARF unwinding for
external calls is implemented by following a link to the current OCaml stack
pointer. With these changes, we get the same backtrace for the \texttt{meander}
program from \S\ref{sec:unwind}, modulo runtime system functions due to effect
handlers. We have verified the correctness of our DWARF directives using the
verification tool from Bastian et al.~\cite{Bastian19}.

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly, which speeds up DWARF-based unwinding by up to 25$\times$. With this
technique, |perf| can unwind the stack at sample points rather than dumping the
call stack, which would capture the complete backtrace rather than just the
current fiber.

\vspace{-1mm}
\subsection{Garbage collection}
\label{sec:gc}

@@

The challenges and the solutions for integrating fibers with the concurrent
mark-and-sweep GC of Multicore OCaml have been discussed
previously~\cite{Sivaramakrishnan20}.

%\vspace{-1mm}
\section{Evaluation}
\label{sec:eval}

In this section, we evaluate the performance of Multicore OCaml effect handlers
against the performance requirements set in \S\ref{sec:req}. Multicore OCaml is
an extension of the OCaml 4.10.0 compiler with support for shared memory
parallelism and effect handlers. Since our objective is to evaluate the impact
of effect handlers, none of our benchmarks utilise parallelism. These results
were obtained on a 2-socket Intel\textregistered Xeon\textregistered Gold 5120
x86-64~\cite{IntelXeonGold5120Spec} server running Ubuntu 18.04 with 64GB of
main memory.

\vspace{-1mm}
\subsection{No effects benchmarks}

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation
of Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation
and re-allocation of resources every time the non-local control leaves and
enters back into a context. |dynamic-wind| is not quite the right abstraction
as resources need to be cleaned up only on non-returning
exits~\cite{Kiselyov,Sitaram03}. This requires distinguishing returning exits
from non-returning ones.
